---
title: "Fitting Stochastic Model on Genetic Data (simulation examples I)"
author: "Mingwei Tang"
date: "`r Sys.Date()`"
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`LNAPhyloDyn` is a R-package that aims at fitting stochastic SIR model using genealogy data. The method and the model are described in the paper ... . Here's an example based on simulated dataset. First, you need to load the package and some other dependent packages. 

```{r,echo=TRUE,message=FALSE}
library(LNAPhyloDyn)
library(MASS)
```


# Simulate data
Here's the parameters for simulation. 

- Initial reproduction number trajectory: R_0 = 2.2
- Recovery rate: gamma = 0.2
- Initial number of infected: I_0 = 1
- Changepoints: At time: 20, with with ratio 1. (ratio =1 means there's at time 20, there's no change in R_0)

Other configurations:

- Time period: [0,100]
- Sampled times: samp_times = c(0,10,20,40,80,85)
- Number of lineages sampled at each time: c(200,200,300,300,20,2)

## Simulate SIR trajectory 

Simulate the infectious SIR trajectory based on SIR model using Markov jump process (MJP): 


```{r,fig.show='hold',fig.height=4,fig.width=7}
x_i4 = c(1,2,0,1) # 1 changepoint, two parameters. 0 is the index for R_0, 1 is the index for gamma. (default for SIR model)
set.seed(44)

traj6 = ODE_rk45(c(100000,1),t = seq(0,100,length.out = 4001), param = c(2.2,0.2,1),x_r = c(100000,20), x_i = x_i4, model = "SIR") # Simulation based on ode

LNA_Traj6 = Traj_sim_ezG2(c(100000,1),times = seq(0,100,length.out = 4001), 
                          param = c(2.2,0.2,1), 40,x_r = c(100000,20), x_i = x_i4,
                          90, model = "SIR")$Simu # simulation based on non-restarting LNA


MJP_Traj6 = simuSIRt(c(100000,1),time = seq(0,100,length.out = 4001), param1 = c(2.2,0.2,1), x_r1 = c(100000,20), x_i1 = x_i4) # simulation based on MJP

# view the simulated data
head(MJP_Traj6)
tail(MJP_Traj6)
# plot S, I trajectory
par(mar = c(4,4,1,1), mgp = c(2,1,0))
plot(MJP_Traj6[,1], MJP_Traj6[,2], type = "l",ylab = "counts", xlab = "time", lwd = 2, col = "red",ylim = c(0,100000))
lines(MJP_Traj6[,1], MJP_Traj6[,3], lwd = 2, col = "blue")
legend("topright", legend = c("Susceptible", "Infected"), lty = 1, lwd =2 , col = c("blue", "red"))
```

## Simulate Genealogies
Given the simulated trajectory `MJP_Traj6`, we simulated the geonealy based on Volz .el 2009. The simulated data contains the sufficient statistics for inference: coalescent time, the sampling time, and the number of lineages sampled. If you want to sample a tree structure, view package `phydynR` by Volz for more details. 

```{r genealogy}
set.seed(55)
coal_simu6 = volz_thin_sir(MJP_Traj6, 90, samp_times = c(0,10,20,40,80,85),
                           c(200,200,300,300,20,2), 
                           betaN = betaTs(c(2.2,0.2,1),t = MJP_Traj6[,1], 
                                          x_r = c(100000,20),x_i = x_i4))
```

### Inference
Now we do the inference by running our MCMC algorithm. First we specify the LNA grid for Ode solver, the LNA grid and the grid for changepoints 

```{r, timeconfig}
times = seq(0,1000,length.out = 2001)
t2 = times[seq(1,length(times),by=50)]
cht = t2[2:(length(t2)-1)]
```


## Start MCMC for 1000 iterations

```{r MCMC,results='hide', message=FALSE, warning=FALSE, cache=TRUE}
set.seed(250)
resres6 = General_MCMC2(coal_simu6, times = seq(0,100,length.out = 2001), t_correct = 90,
                        N = 100000,gridsize = 50,niter = 1000,burn = 0,thin = 5,
                        changetime = cht,DEMS = c("S","I"),
                        prior=list(pop_pr=c(1,1,1,1), R0_pr=c(1,7), gamma_pr = c(-1.5,0.1), mu_pr = c(-1.2,0.1), hyper_pr=c(30,0.7)),
                        proposal = list(pop_prop = 0.2, R0_prop = c(0.01), gamma_prop=0.05, mu_prop = 0.02, hyper_prop=0.25),
                        control = list(alpha=c(1/100000), R0 = 1.8, gamma=0.2, ch=rep(1,length(cht)), traj = matrix(rep(0,2*40),ncol=2),hyper = 10),updateVec = c(1,1,1,0,0.5,1,1),
                        likelihood = "volz", model = "SIR",
                        Index = c(0,1), nparam=2,method = "admcmc", options=list(joint = T, PCOV = NULL,beta=0.95,burn1 = 10, parIdlist = list(a = 2,b=3,c=4,d=5 + length(cht)), isLoglist = list(a = 1,b=0,c=1,d=0),
                                                                                 priorIdlist = list(a=1,b=2,c = 3,d = 5),up = 1000000), verbose = F)

```

### Simulation results
```{r show sim_res, fig.show='hold',fig.height=3, fig.width=5, cache=TRUE,dependson = 'MCMC'}
# R0(t) trajectory 
par(mar = c(4,4,1,1), mgp = c(2,1,0))
randomR0_traj(seq(0,100,length.out = 41),resres6,3,c(5:43),
              seq(100,1000,by=5),ylim = c(1,3))
abline(h = 2.2,col = "red", lwd = 2)
# Histogram for recovery rate gamma
hist(resres6$par[100:1000,4],main = "", xlab = "gamma")
vlineCI(resres6$par[100:1000,4])
abline(v = 0.2, col = "red", lwd = 2)
```

If you want to have better results, try more iterations. 

Have fun! 

<!--Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
-->
## Reference

1. Volz E M, Pond S L K, Ward M J, et al. Phylodynamics of infectious disease epidemics[J]. Genetics, 2009, 183(4): 1421-1430.
